Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-27 manuscript/draft.Rmd
manuscript stuff

---
title: Using best practices for forecasting biodiversity changes in breeding birds
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
bibliography: refs.bib
geometry: margin=1in
fontsize: 12pt
header-includes:
    - \usepackage{times}
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
output:
  pdf_document: default
  html_document: default
---

`r settings = yaml::yaml.load_file("../settings.yaml")`

@@

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
to quantify the spatial relationships between the occurrence of individual
species and the environment [@thomas2004; @urban2015]. Forecasts
for future environmental conditions are then used to predict where each species
will occur in the future and the probabilities of occurrence are summed across
species to predict future species richness [e.g., @thuiller2005]. Alternatively,
models that directly relate spatial patterns of species richness to the
environment have been developed and generally perform equivalently to species
distribution modeling based methods [@algar2009; @distler2015].

However, despite the emerging interest in forecasting species richness and other
aspects of biodiversity, little is known about how effectively we can anticipate
their dynamics. This is due in part to the long time scales over which many
ecological forecasts are applied and the resulting difficulty in assessing
whether the predicted changes occurred [@dietze2016]. What we do know comes
from a small number of hindcasting studies, where models are built using data on
species occurrence and richness from the past and evaluated on their ability to
predict contemporary patterns [e.g., @algar2009; @distler2015]. These studies
are a valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation
[@clark2001, @dietze2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine if the forecasts are more accurate than
assuming the system is basically static [@perretti2013].

In this paper we attempt to forecast the species richness of breeding birds at
hundreds of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this we combine 30 years of
time-series data on bird distributions with monthly time-series of climate data
and satellite-based remote-sensing. This 30 year annual time-scale has only
recently become possible for large-scale time-series based forecasting and
allows us to model and assess changes a decade or more into the future in the
presence of shifts in environmental conditions on par with predicted climate
change. We compare traditional distribution modeling based approaches to spatial
models of species richness, time-series methods, and simple baselines. All
forecasts model uncertainty and observation, are evaluated across different time
lags using hindcasting, and are publicly archived to allow future assessment. We
discuss the implications of these practices for our understanding of, and
confidence in, the resulting forecasts, and how we can continue to build on
these approaches improve ecological forecasting in the future.

@@

We used observed richness values from 1982 (the first complete year of NDVI
data) to 2003 for training the models, and data from 2004 to 2013 was used for
testing their performance. We only used BBS routes from the continental United
States (i.e. routes where PRISM data was available), and restricted the analysis
to routes that were sampled during 70% of the years in the training period
(i.e., routes with at least 16 annual observations). The resulting dataset
included [[N]] transect runs of [[N]] unique sites, and included [[N]] species.
Site-level richness varied from [[N]] to [[N]] with an average richness of [[N]]
species.

@@

Observer effects are inherent in large data sets taken by different observers,
and are known to occur in BBS [@sauer1994]. We estimated the observer effects
(and associated uncertainty about those effects) with a linear mixed model with
observer as a random effect built in the Stan probabilistic programming language
[@stan] and its R interface [@rstan]. Because observer and site are strongly
related (an observer tends to repeatedly sample the same site), site was also
included as a random effect to ensure that inferred deviations were actually
observer-related (as opposed to being related to the sites that a given observer
happened to see). The resulting model partitions the variance in observed
richness values into site-level variance, observer-level variance, and residual
variance (e.g. variation within a site from year to year), and can therefore
also be used directly as the "average" baseline model (see below). The estimated
observer effects can be subtracted from the richness values for a particular
observer to provide an estimate of how many species would have been found by a
"typical" observer. To incorporate uncertainty in these "corrected" richness
values into the forecasting models we collected 500 Monte Carlo samples from the
model's posterior distribution, and fit each of the downstream models with each
of the Monte Carlo samples.

@@

We fit three sets of models that were fit to each site independently. These
models were fit to each BBS route both on the raw richness values and on the
residuals from our observer model. When correcting for observer effects, we
averaged across 500 models that were fit separately to the 500 Monte Carlo
estimates of the observer effects, to account for our uncertainty in the true
values of those effects. All of these models use a Gaussian error distribution.

**Baseline models.** We used two simple models as baselines for comparison to
the fits of more complex models. These baselines treated site-level richness
observations either as uncorrelated noise around a site-level constant (the
"average" model) or as an autoregressive model with a single year of history
(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
centered on the average richness observed during training, and the confidence
intervals are narrow and constant-width. The "naive" model, in contrast,
predicts that future observations will be similar to the final observed value
(e.g. the value observed in `$r settings$last_train_year`), and the confidence
intervals expand rapidly as the predictions extend farther into the future.

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components, and how many terms to include for each one,
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function.

@@

In contrast to the time-series models, most attempts to predict species richness
focusing on using correlative models to predict richness changes over time from
environmental data. We tested three common variants of this approach - direct
modeling of species richness, stacking individual species distribution models,
and joint species distribution models (JSDMs). Following the standard approach,
site-level random effects were not included in these models as predictors,
meaning that this approach implicitly assumes that two sites with identical
Bioclim, elevation, and NDVI values should have identical richness
distributions. As above, we included observer effects & the associated
uncertainty by running these models 500 times (once per MCMC sample).

**"Macroecological" model.** We used a boosted regression tree model using the
`gbm` package [@ridgeway2017] to directly model species richness as a function
of environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its predecessors.
They are flexible models that are considered well suited for prediction
[@elith2008]. This model was optimized using a Gaussian likelihood, with a
maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees.
The number of trees was selected using the "out of bag" estimator, and
[[summarize how many trees were typically chosen here, for both observer-based
and observer-free models]].

**Species Distribution Model.** Species distribution models (SDMs) predict
individual species' occurrence probabilities using environmental variables.
Species-level models are used to predict richness by summing the predicted
probability of occupancy across all species at a site; following [[Calabrese et
al.]], we calculated the uncertainty in our richness estimate by treating
richness as a sum over independent Bernoulli random variables (one per species).
This avoids known problems with the use of thresholds for determining whether or
not a species will be present at a site [[refs]]. We represented SDMs in this
analysis using random forests, a common approach in the species distribution
modeling literature based on fitting hundreds of independent regression trees to
randomly-perturbed versions of the data [[refs; would be good to provide a
source that says these are a good choice]]. Each of the 500 trees in our
species-level random forests used a different Monte Carlo estimate of the
observer effects as a predictor variable.

@@

Variance in species richness within sites was lower than predicted by several
common count models, such as the Poisson or binomial (i.e. richness was
underdispersed), so these count models would have had difficulty fitting the
data. Since richness had a relatively large mean ([[N]]) and all models already
produce continuous richness estimates (due to summing probabilities in the
(J)SDM case), we defined model performance for all models in terms of continuous
Gaussian errors instead of using discrete count distributions. When a model was
run multiple times for the purpose of correcting for observer effects, we used
the mean of those runs' point estimates as our final point estimate and we
calculated the uncertainty using the law of total variance (i.e.  the average of
the model runs' variance, plus the variance in the point estimates).

We evaluated the performance of each model for forecasting using the data for
each year between `r settings$last_train_year` and `r settings$end_yr`. We used
three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
determine how far, on average, the models' predictions were from the observed
value; 2) the 95% prediction interval coverage to determine how well the models
predicted the range of possible outcomes; and 3) deviance (i.e. negative 2 times
the Gaussian log-likelihood) as an integrative measure of fit incorporating good
point estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon.

All analyses were conducted using R [@R]. Primary R packages used in the
analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp
[@sp1; @sp2], raster [@raster], prism [@prism], rdataretriever
[@rdataretriever], forecasts [@forecast1; @forecast2], git2r [@git2r], ggplot
[@ggplot2], viridius [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
gbm [@gbm], randomForest [@randomForest], purrrlyr [@purrrlyr]. Code to fully
reproduce this analysis is available on GitHub
(https://github.com/weecology/bbs-forecasting) and archived on Zenodo ().

>>>>

---
title: Using best practices for forecasting biodiversity changes in breeding birds
author:
- David J. Harris
- Shawn D. Taylor
- Ethan P. White
bibliography: refs.bib
csl: csl.csl
geometry: margin=1in
fontsize: 12pt
header-includes:
    - \usepackage{times}
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
output:
  pdf_document: default
  html_document: default
---

`r settings = yaml::yaml.load_file("../settings.yaml")`
`r timeframe = settings$timeframes$train_22`
`r numbers = yaml::yaml.load_file("numbers.yaml")`

@@

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models.
In general, these models describe individual species' occurrence patterns across
sites as functions of the site's environment [@thomas2004; @urban2015]. By
plugging forecasts of future environmental conditions into these functions,
ecologists can make predictions about where each species will occur in the
future. These species-level predictions can then be added up ("stacked") to
generate predicted richness values [e.g. @calabrese2014]. Alternatively, models
that directly relate spatial patterns of species richness to the environment
have been developed and generally perform equivalently to species distribution
modeling based methods [@algar2009; @distler2015]. This approach is sometimes
referred to as "macroecological" modeling, because it models the larger-scale
pattern (richness) more directly [@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity, little is known about how effectively we can anticipate their
dynamics. This is due in part to the long time scales over which many ecological
forecasts are applied and the resulting difficulty in assessing whether the
predicted changes occurred [@dietze2016]. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009; @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation (Box
1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation [@clark2001,
@dietze2017]; and 3) conducting meaningful evaluations of the forecasts by
hindcasting, archiving short-term forecasts, and comparing forecasts to
baselines to determine whether the forecasts are more accurate than simply
assuming the system is basically static [@perretti2013].

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. This 30 year
time-scale has only recently become possible for large-scale time-series based
forecasting and allows us to model and assess changes a decade or more into the
future in the presence of shifts in environmental conditions on par with
predicted climate change. We compare traditional distribution modeling based
approaches to spatial models of species richness, time-series methods, and
simple baselines. All forecasts model uncertainty and observation, are evaluated
across different time lags using hindcasting, and are publicly archived to allow
future assessment. We discuss the implications of these practices for our
understanding of, and confidence in, the resulting forecasts, and how we can
continue to build on these approaches improve ecological forecasting in the
future.

@@

We used observed richness values from 1982 (the first complete year of NDVI
data) to 2003 for training the models, and data from 2004 to 2013 was used for
testing their performance. We only used BBS routes from the continental United
States (i.e. routes where PRISM data was available), and restricted the analysis
to routes that were sampled during 70% of the years in the training period
(i.e., routes with at least 16 annual observations). The resulting dataset
included `r numbers$N_runs` transect runs of `r numbers$N_sites` unique sites,
and included `r numbers$N_species` species. Site-level richness varied from `r
numbers$richness_summary$min` to `r numbers$richness_summary$max` with an
average richness of `r numbers$richness_summary$mean` species.

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan] and its R interface [@rstan]. Because observer and
site are strongly related (observers tend to repeatedly sample the same site),
site was also included as a random effect to ensure that inferred deviations
were actually observer-related (as opposed to being related to the sites that a
given observer happened to see). The resulting model partitions the variance in
observed richness values into site-level variance, observer-level variance, and
residual variance (e.g. variation within a site from year to year). The
site-level estimates can also be used directly as the "average" baseline model
(see below). The estimated observer effects can be subtracted from the richness
values for a particular observer to provide an estimate of how many species
would have been found by a "typical" observer. To incorporate uncertainty in
these "corrected" richness values into the forecasting models we collected 500
Monte Carlo samples from the model's posterior distribution, and fit each of the
downstream models with each of the Monte Carlo samples.

@@

We fit three sets of models that were fit to each site independently, with no
environmental information. These models were fit to each BBS route both on the
raw richness values and on the residuals from our observer model. When
correcting for observer effects, we averaged across 500 models that were fit
separately to the 500 Monte Carlo estimates of the observer effects, to account
for our uncertainty in the true values of those effects. All of these models use
a Gaussian error distribution.

**Baseline models.** We used two simple models as baselines for comparison to
the fits of more complex models. These baselines treated site-level richness
observations either as uncorrelated noise around a site-level constant (the
"average" model) or as an autoregressive model with a single year of history
(the "naive" model) [@hyndman2014]. Predictions from the "average" model are
centered on the average richness observed during training, and the confidence
intervals are narrow and constant-width. The "naive" model, in contrast,
predicts that future observations will be similar to the final observed value
(e.g. the value observed in `r timeframe$last_train_year`), and the confidence
intervals expand rapidly as the predictions extend farther into the future.

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components (and how many terms to include for each one)
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function.

@@

In contrast to the time-series models, most attempts to predict species richness
focus on using correlative models to predict richness changes over time from
environmental data. We tested three common variants of this approach - direct
modeling of species richness, stacking individual species distribution models,
and joint species distribution models (JSDMs). Following the standard approach,
site-level random effects were not included in these models as predictors,
meaning that this approach implicitly assumes that two sites with identical
Bioclim, elevation, and NDVI values should have identical richness
distributions. As above, we included observer effects & the associated
uncertainty by running these models 500 times (once per MCMC sample).

**"Macroecological" model.** We used a boosted regression tree model using the
`gbm` package [@ridgeway2017] to directly model species richness as a function
of environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its predecessors.
They are flexible models that are considered well suited for prediction
[@elith2008]. This model was optimized using a Gaussian likelihood, with a
maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees. The
number of trees used for prediction was selected using the "out of bag"
estimator; this number averaged [[N]] for the non-observer data and [[N]] for
the observer-corrected data.

**Species Distribution Model.** Species distribution models (SDMs) predict
individual species' occurrence probabilities using environmental variables.
Species-level models are used to predict richness by summing the predicted
probability of occupancy across all species at a site. This avoids known
problems with the use of thresholds for determining whether or not a species
will be present at a site [@calabrese2014; pellissier2013]. Following
@calabrese2014, we calculated the uncertainty in our richness estimate by
treating richness as a sum over independent Bernoulli random variables:
$\sigma^2_{richness} = \sum_i{p_i (1 - p_i)}$, where $i$ indexes species. By
itself, this approach is known to underestimate the true community-level
uncertainty because it ignores the uncertainty in the species-level probabilites
[@calabrese2014]. To mitigate this problem, we used an ensemble of 500 estimates
for each of the species-level probabilities instead of just one, propagating the
uncertainty forward. We obtained these estimates using using random forests, a
common approach in the species distribution modeling literature based on fitting
hundreds of independent regression trees to randomly-perturbed versions of the
data [@cutler2007; @caruana2008]. When correcting for observer effects, each of
the 500 trees in our species-level random forests used a different Monte Carlo
estimate of the observer effects as a predictor variable.

@@

Variance in species richness within sites was lower than predicted by several
common count models, such as the Poisson or binomial (i.e. richness was
underdispersed for individual sites), so these count models would have had
difficulty fitting the data. Since richness had a relatively large mean ([[N]])
and all models already produce continuous richness estimates (due to summing
probabilities in the (J)SDM case), we defined model performance for all models
in terms of continuous Gaussian errors instead of using discrete count
distributions. When a model was run multiple times for the purpose of correcting
for observer effects, we used the mean of those runs' point estimates as our
final point estimate and we calculated the uncertainty using the law of total
variance (i.e.  the average of the model runs' variance, plus the variance in
the point estimates).

We evaluated the performance of each model for forecasting using the data for
each year between `r timeframe$last_train_year` and `r timeframe$end_yr`. We used
three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
determine how far, on average, the models' predictions were from the observed
value; 2) the 95% prediction interval coverage to determine how well the models
predicted the range of possible outcomes; and 3) deviance (i.e. negative 2 times
the Gaussian log-likelihood) as an integrative measure of fit incorporating good
point estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon.

All analyses were conducted using R [@R]. Primary R packages used in the
analysis included dplyr [@dplyr], tidyr [@tidyr], gimms [@gimms], sp [@sp1;
@sp2], raster [@raster], prism [@prism], rdataretriever [@rdataretriever],
forecast [@forecast1; @forecast2], git2r [@git2r], ggplot [@ggplot2], mistnet
[@harris2015], viridis [@viridis], rstan [@rstan], yaml [@yaml], purrr [@purrr],
gbm [@gbm], randomForest [@randomForest]. Code to fully reproduce this analysis
is available on GitHub (https://github.com/weecology/bbs-forecasting) and
archived on Zenodo ().
